---
title: "BDA_2024_MetaAnalysis_Glioblastoma"
author: "Duy Nguyen"
date: "2024-06-25"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```

### 1) Project Set Up

```{r setdw, eval = FALSE, echo = FALSE}
# Set working directory 
setdw("/Users/manhduynguyen/BDA_2024_Glioblastoma")
```


```{r renv, eval = FALSE, echo = FALSE}
# Initiate the project with renv
renv::init()

# Save to lockfile
renv::snapshot()
```


```{r install_pack, eval = FALSE, echo = FALSE}
# Download packages
if(!requireNamespace("devtools", quietly = T)){
  install.packages("devtools")
}
devtools:: install_github("PavlidisLab/gemma.R", force = T)
install.packages("plyr")
install.packages("tidyverse")
```


```{r load_pack}
# Load packages
library("gemma.R")
library("plyr")
library("tidyverse")

# Check loaded packages 
(.packages())
```


### 2) Gemma Dataset Search

- Datasets in Gemma are annotated using structured terms of ontologies. Many of
the ontological terms are hierarchical.

- **`Ontologies`**: a set of concepts and categories in a subject area or domain 
that shows their properties and the relations between them.

- Within an ontological hierarchy, the broader, umbrella term is called the 
*"parent"* while the term representing the more specific subset is called the 
*"child"*. When requesting Gemma records with particular annotation, a parent 
term can often pull up all records with the child terms.


```{r onto_anno, cache = TRUE}
# Search for ontological annotations of "glioblastoma"
gbm_annotation <- search_annotations("glioblastom*")

# Count usage of ontological terms and filter terms that are used in Gemma
gbm_annotation_in_Gemma <- gbm_annotation %>%
  # Add new column "counts" in which the function is being applied to every
  # single value.URI from gbm_annotation
  mutate(counts = sapply(value.URI, function(uri) {
    # This function retrieves "totalElements" attribute from the datasets 
    # associated with given input URI
    # The "totalElements" represent the total number of records
    attributes(get_datasets(uris = uri))$totalElements
  })) %>%
  # Filter to keep URI with at least 1 records
  filter(counts > 0)

# Check child terms for "glioblastoma"
get_child_terms("http://purl.obolibrary.org/obo/MONDO_0018177")
```


```{r search_data, cache = TRUE}
# Search dataset using search terms
gbm_datasets_searchterm <- get_datasets(query = "glioblastom* OR 
                                                gbm OR gb OR 
                                                \"glioblastoma multiforme\"
                                                \"grade IV astrocytic tumor\" OR 
                                                \"grade IV astrocytoma\" OR 
                                                \"WHO grade IV glioma)\"",
                                        # Focus on human datasets
                                        taxa = c("human")) %>%
  # Get full set of results
  get_all_pages()

# Search dataset using parent term on ontological hierarchy
gbm_datasets_uris <- get_datasets(uris = "http://purl.obolibrary.org/obo/MONDO_0018177",
                                  # Focus on human datasets
                                  taxa = c("human")) %>%
  # Get full set of results
  get_all_pages()

# Check if the datasets are duplicated in both methods of searching
gbm_datasets_check <- gbm_datasets_searchterm %>%
  mutate(available_in_both_dataframes = ifelse(experiment.shortName %in% 
                                               gbm_datasets_uris$experiment.shortName, 
                                               TRUE, FALSE))


# Check the number of datasets
length(gbm_datasets_check$taxon.name)

# Check the quality of the datasets
# (FALSE - data are of good quality)
table(gbm_datasets_check$experiment.troubled)

# Check if raw data were available 
# (1 - raw data available | -1 - raw data not available)
table(gbm_datasets_check$experiment.rawData)

# Filter for datasets with good quality and raw data available
gbm_datasets_filtered <- gbm_datasets_check %>%
  filter(experiment.troubled == FALSE & gbm_datasets_check$experiment.rawData == 1)
```







